Individual and Differential Harm in Redistricting

### 2022-08-01 paper/harm.Rmd
minivation

# Motivation {#sec:motivation}

To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representations of Missouri.
Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority.
Minissouri has a population of 180 and should be divided into 3 districts with population between 54 and 66.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
A schematic for this example is shown in Figure \ref{fig:minissouri}.

## Minichusetts: counterfactuals matter

Every precinct in Minichusetts has a Democratic majority.
As a result, it is impossible to draw a districting plan built from precincts that creates even a single Republican district.
This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].

This Democratic dominance poses a problem for existing measures of partisan gerrymandering.
One of the most common measures, the efficiency gap of @stephanopoulos2015, takes a value of $-0.311$ or $-0.317$ for Minichusetts, depending on the configuration of the districts.
This would make any Minichusetts districting plan an extreme Democratic gerrymander in comparison to the historical range of efficiency gaps observed in states nationwide.[^1]

[^1]: See, e.g., the historical distribution calculated by the Campaign Legal Center's PlanScore tool: <https://planscore.campaignlegal.org/metrics/efficiencygap/>.

But it is impossible to gerrymander if every manipulation of district boundaries produces no change in partisan outcomes.
What the efficiency gap and other gerrymandering measures lack is an awareness of the space of possible districting plans, plans that *counterfactually* might have been selected.
Once a set of districting rules has been agreed on (in Minichusetts, that districts must be contiguous and composed of whole precincts), what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
There can be no gerrymander, and no harm to individual voters, if no improvement in partisan fairness is possible.

## Mininois: individual harm

To explore how to properly integrate counterfactual districting plans into the analysis of a potential gerrymander, we turn to the more interesting state of Mininois, where, in a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].

Consider the two plans in Figure XX(a)-(b).
A Democratic gerrymander (a) effectuates Democratic control by splitting its city, pinwheel-like, into three districts reaching out into the more Republican and sparsely populated remainder of the state.
This gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.

Suppose now that plan (a) is enacted into law.
Who does this decision harm, if anyone?
Precisely those voters who are not represented by their chosen party, but would have been counterfactually.
Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
Republicans in the city and in the southwest are unharmed, as are all Democratic voters, since they are represented by the same party under both plans.
Figure XX, Panel C plots the number of voters harmed in each precinct.

Measuring the harm of the Democratic gerrymander required comparing it to a neutral counterfactual plan.
The choice of counterfactual might seem hopelessly subjective.
We discuss in Section \ref{sec:estimating} principled ways to define counterfactuals, but for now we hope the value in measuring harm through counterfactuals is clear even if the operationalization is not.

## Minissouri: differential harm

We could quantify the harm a Democratic gerrymander imposed on voters in Mininois by comparing election outcomes to those under a counterfactual proportional plan.
But the results said nothing about the direction of the harm.
While it is apparent that the Democratic gerrymander indeed benefits Democrats, tallying up the total harm done to all voters is not enough to prove this.

However, we can partition voters into groups, and compare how the harm done by a plan falls across those groups.
If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
Figure \ref{fig:minissouri} illustrates this idea for two hypothetical gerrymanders ((a) and (c)) of Minissouri.
As in Mininois, we can tally the total number of voters harmed in each precinct ((d) and (f)) in comparison to a proportional counterfactual plan (b).
In both cases, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

@@

Panel (e) then splits the voters by party, and computes, for each party, the share of its voters which are harmed.
The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats, and the opposite can be said of the Republican gerrymander (c).
We call the gap between these two rates the *differential harm,* and propose it as a general way to account for the impact of a districting plan on particular groups.

The differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander is $0.450$, where positive values indicate Republican bias.
In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan in Figure \ref{fig:minissouri}(b).

>>>>

# Minivation {#sec:motivation}

To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representations of Missouri.
Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority.
Minissouri has a population of 180 and should be divided into 3 districts with population between 54 and 66.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
A schematic for this example is shown in Figure \ref{fig:minissouri}.
Each person is a square or circle, identifying some group membership or characteristic.
Each person is then shaded red or blue to correspond to Republicans and Democrats, respectively.

@@

Now, consider the two plans in Figure \ref{fig:minissouri}(a)-(b).
Like in much of the US, in Minissouri Democrats are sorted asymmetrically into the cities while Republicans dominate the rural regions [@chen:rodd:13; @enos:brown:21].
A Democratic gerrymander (a) effectuates Democratic control by splitting its cities, connecting them through thin regions around the boundary of the state.
This gerrymander would elect one Republican and two Democrats, whereas the proportional map (b) would elect two Republicans and one Democrat.

If we take the Democratic gerrymander as the reference plan and the proportional plan as the counterfactual, Figure \ref{fig:minissouri}(c) highlights who is harmed.
Most of the highlighted people are the Republicans who were placed in a Democratic district in the gerrymander, but are now in a Republican district in the proportional plan.
Note that the Republicans in the north-east city are not harmed here.
In both plans, those Republicans are in a Democratic district in both the reference and the counterfactual.
There are two blue people who are harmed, as they were placed into a Republican district in the gerrymander, but now move to a Democratic district.
We can say then that 44 Republicans and 2 Democrats are individually harmed.

There's a second dimension here to individual harm.
Among the people, some are circles and some are squares.
Squares are highly correlated with the Republican choice as circles are with the Democratic choice.
Based on their vote choice, as before, we can identify who is harmed.
Now, there are circles harmed from both parties.
2 blue circles, 11 red circles, 33 red squares, and no blue squares are individually harmed.

We can use these groups, to compare how the harm done by a plan falls across those groups.
If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.

The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats.
We call the gap between these two rates the *differential harm,* and propose it as a general way to account for the impact of a districting plan on particular groups.
If there is no gap, then we have *fairness*.

The partisan differential harm of the Democratic gerrymander of Minissouri is $-0.415$, where positive values indicate Republican bias.
In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan.
The shape-based differential harm of the gerrymander is $-0.2$, where positive values indicate pro-square bias.
On account of being a square and not a circle, a square is 20% less likely to be represented by their candidate of choice.
